---
title: "PDA Project"
author: "Andrey"
date: "2024-03-19"
output:
  pdf_document: default
  html_document: default
---


```{r}
# Libraries
library(ggplot2)
library(Hmisc)
library(tidyr)
library(dplyr)
library(reshape2)
library(corrplot)
library(ggcorrplot)
library(factoextra)
library(Metrics)
library(validate)
library(stats)
library(fpc)
library(cluster)
```

```{r}
read.csv("WHO.csv")
df <- read.csv("WHO.csv")
```

```{r}
# Checking the summary of the data
summary(df)
# Checking the structure of the data
str(df)
```

```{r}
# Checking the missing values
colSums(is.na(df))
```

```{r}
# Calculating the number of outliers in the numeric columns
outliers <- function(data) {
  Q1 <- quantile(data, 0.25, na.rm = TRUE)
  Q3 <- quantile(data, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  outliers <- data[data < lower_bound | data > upper_bound & !is.na(data)]
  return(outliers)
}

IQR_year <- outliers(df$Year)
IQR_Life.expectancy <- outliers(df$Life.expectancy)
IQR_Unemployment <- outliers(df$Unemployment)
IQR_Infant.Mortality <- outliers(df$Infant.Mortality)
IQR_GDP <- outliers(df$GDP)
IQR_GNI <- outliers(df$GNI)
IQR_Clean.fuels.and.cooking.technologies <- outliers(df$Clean.fuels.and.cooking.technologies)
IQR_Per.Capita <- outliers(df$Per.Capita)
IQR_Mortality.caused.by.road.traffic.injury <- outliers(df$Mortality.caused.by.road.traffic.injury)
IQR_Tuberculosis.Incidence <- outliers(df$Tuberculosis.Incidence)
IQR_DPT.Immunization <- outliers(df$DPT.Immunization)
IQR_HepB3.Immunization <- outliers(df$HepB3.Immunization)
IQR_Measles.Immunization <- outliers(df$Measles.Immunization)
IQR_Hospital.beds <- outliers(df$Hospital.beds)
IQR_Basic.sanitation.services <- outliers(df$Basic.sanitation.services)
IQR_Tuberculosis.treatment <- outliers(df$Tuberculosis.treatment)
IQR_Urban.population <- outliers(df$Urban.population)
IQR_Rural.population <- outliers(df$Rural.population)
IQR_Non.communicable.Mortality <- outliers(df$Non.communicable.Mortality)
IQR_Sucide.Rate <- outliers(df$Sucide.Rate)

print(paste("Number of outliers in Year:", length(IQR_year)))
print(paste("Number of outliers in Life.expectancy:", length(IQR_Life.expectancy)))
print(paste("Number of outliers in Unemployment:", length(IQR_Unemployment)))
print(paste("Number of outliers in Infant.Mortality:", length(IQR_Infant.Mortality)))
print(paste("Number of outliers in GDP:", length(IQR_GDP)))
print(paste("Number of outliers in GNI:", length(IQR_GNI)))
print(paste("Number of outliers in Clean.fuels.and.cooking.technologies:", length(IQR_Clean.fuels.and.cooking.technologies)))
print(paste("Number of outliers in Per.Capita:", length(IQR_Per.Capita)))
print(paste("Number of outliers in Mortality.caused.by.road.traffic.injury:", length(IQR_Mortality.caused.by.road.traffic.injury)))
print(paste("Number of outliers in Tuberculosis.Incidence:", length(IQR_Tuberculosis.Incidence)))
print(paste("Number of outliers in DPT.Immunization:", length(IQR_DPT.Immunization)))
print(paste("Number of outliers in HepB3.Immunization:", length(IQR_HepB3.Immunization)))
print(paste("Number of outliers in Measles.Immunization:", length(IQR_Measles.Immunization)))
print(paste("Number of outliers in Hospital.beds:", length(IQR_Hospital.beds)))
print(paste("Number of outliers in Basic.sanitation.services:", length(IQR_Basic.sanitation.services)))
print(paste("Number of outliers in Tuberculosis.treatment:", length(IQR_Tuberculosis.treatment)))
print(paste("Number of outliers in Urban.population:", length(IQR_Urban.population)))
print(paste("Number of outliers in Rural.population:", length(IQR_Rural.population)))
print(paste("Number of outliers in Non.communicable.Mortality:", length(IQR_Non.communicable.Mortality)))
print(paste("Number of outliers in Sucide.Rate:", length(IQR_Sucide.Rate)))
```
```{r}
#Check any misspell in country names
table(df$Country)
```
```{r}
#comprehension checking for each variables using a set of validation rules. 
life_data_check <- validator(
      
      #Check if there is any duplicated rows
      duplicated_rows = anyDuplicated(df),
      #No negative value in numerical variables
      okYear = Year>=0,
      okLife.expectancy =Life.expectancy >=0,
      okUnemployment = Unemployment >= 0 & Unemployment <= 100,
      okInfant.Mortality =Infant.Mortality >= 0& Infant.Mortality <= 1000,
      okGDP = GDP >= 0,
      okGNI = GNI >= 0,
      okClean.fuels.and.cooking.technologies = Clean.fuels.and.cooking.technologies >= 0& Clean.fuels.and.cooking.technologies <= 100,
      okPer.Capita = Per.Capita >= 0,
      okMortality.caused.by.road.traffic.injury = Mortality.caused.by.road.traffic.injury >= 0& Mortality.caused.by.road.traffic.injury <= 100000,
      okTuberculosis.Incidence = Tuberculosis.Incidence >= 0& Tuberculosis.Incidence <= 100000,
      okDPT.Immunization = DPT.Immunization >= 0& DPT.Immunization <= 100,
      okHepB3.Immunization = HepB3.Immunization >= 0& HepB3.Immunization <= 100,
      okMeasles.Immunization = Measles.Immunization >= 0& Measles.Immunization <= 100,
      okHospital.beds = Hospital.beds >= 0& Hospital.beds <= 1000,
      okBasic.sanitation.services =Basic.sanitation.services >= 0& Basic.sanitation.services <= 100.00,
      okTuberculosis.treatment = Tuberculosis.treatment >= 0& Tuberculosis.treatment <= 100,
      okUrban.population = Urban.population >= 0& Urban.population <= 100,
      okRural.population = Rural.population >= 0& Rural.population <= 100,
      okNon.communicable.Mortality = Non.communicable.Mortality >= 0& Non.communicable.Mortality <= 1000,
      okSucide.Rate = Sucide.Rate >= 0& Sucide.Rate <= 100000
                  )                     
                      
qual.check<- confront(df, life_data_check)
summary(qual.check)
#Create a validation plot
plot(qual.check)
```
## Data Cleaning

```{r}
# Converting the categorical variables to factors
df$Country <- as.factor(df$Country)
df$Gender <- as.factor(df$Gender)
```

```{r}
# Log transformation of GDP and GNI to reduce skewness
df$GDP <- log(df$GDP)
df$GNI <- log(df$GNI)
```

## EDA
```{r}
#Histogram for numerical variables

num_cols <- df[c("Year", "Life.expectancy", "Unemployment", "Infant.Mortality", "GDP", "GNI", "Clean.fuels.and.cooking.technologies", "Per.Capita", "Mortality.caused.by.road.traffic.injury", "Tuberculosis.Incidence", "DPT.Immunization", "HepB3.Immunization", "Measles.Immunization", "Hospital.beds", "Basic.sanitation.services", "Tuberculosis.treatment", "Urban.population", "Rural.population", "Non.communicable.Mortality", "Sucide.Rate")]

for (col in colnames(num_cols)) {
  histogram_plot <- ggplot(num_cols, aes(x = .data[[col]])) +
                      geom_histogram(bins =20,size = 0.2, fill = 'grey',color = 'black', aes(y = ..density..)) +
                      geom_density(alpha=.2,size = 0.3, fill="blue") +
                      labs(x = col, y = 'Density') +
                      theme_minimal() +
                      ggtitle(paste('Histogram and Density of', col))+
                      theme(panel.border = element_rect(colour = "black", fill=NA, size=1))
  print(histogram_plot)
}
```
```{r}
# Boxplots for each numeric variable

for (col in colnames(num_cols)) {
  boxplot_plot <- ggplot(num_cols, aes(y = .data[[col]], x = 1)) +
                  geom_boxplot(fill = 'grey', size = 0.3 , color = 'black', outlier.color =  "black", outlier.size = 0.5,notch = TRUE)+
                    labs(y = col, x = '') +
                    theme_minimal() +
                    ggtitle(paste('Boxplot of', col))+
                    theme(panel.border = element_rect(colour = "black", fill = NA, size = 1))
  print(boxplot_plot)
}
```

```{r}
# Barplot for Gender
ggplot(df, aes(x = reorder(Gender, -table(Gender)[Gender]))) +
    geom_bar(fill = 'gray', color = 'black', stat = "count") +
    xlab("") + geom_text(stat = "count", aes(label = ..count..), vjust = -0.3) +
    ggtitle("Barplot of Brand") +
    theme(axis.text.x = element_text(hjust = 1))
    
```
```{r}
# Scaterplot for numerical variables vs Life.expectancy

num_cols2 <- c("Year", "Unemployment", "Infant.Mortality", "GDP", "GNI", "Clean.fuels.and.cooking.technologies", "Per.Capita", "Mortality.caused.by.road.traffic.injury", "Tuberculosis.Incidence", "DPT.Immunization", "HepB3.Immunization", "Measles.Immunization", "Hospital.beds", "Basic.sanitation.services", "Tuberculosis.treatment", "Urban.population", "Rural.population", "Non.communicable.Mortality", "Sucide.Rate")

for (var in num_cols2) {
  scatter_plot <- ggplot(df, aes_string(x = var, y = 'Life.expectancy')) +
                    geom_point(alpha = 0.3, size = 0.5, color = "blue") +
                    geom_smooth(color = "darkorange", fill = "darkorange", method = "lm", span = 0.2) +
                    labs(title = paste('Scatter Plot of Life.expectancy vs', var), x = var, y = 'Life.expectancy') +
                    theme_minimal()+
                    theme(panel.border = element_rect(colour = "black", fill=NA, size=0.5),
                    axis.text.x = element_text(angle = 0, hjust = 1))
  print(scatter_plot)
}
```

```{r}
# Boxplot for Gender vs Life expectancy
png("fm.png", width=1200, height=800, res=100)
box_gender <- ggplot(df, aes(x = Gender, y = Life.expectancy , fill = Gender)) +
  geom_boxplot() +
  ggtitle("Life expectancy by Gender")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0, hjust = 1), legend.position = "none")

print(box_gender)
dev.off()
```
```{r}
# Heatmap for correlation matrix

num_cols3 <- df[c("Year","Life.expectancy", "Unemployment", "Infant.Mortality", "GDP", "GNI", "Clean.fuels.and.cooking.technologies", "Per.Capita", "Mortality.caused.by.road.traffic.injury", "Tuberculosis.Incidence", "DPT.Immunization", "HepB3.Immunization", "Measles.Immunization", "Hospital.beds", "Basic.sanitation.services", "Tuberculosis.treatment", "Urban.population", "Rural.population", "Non.communicable.Mortality", "Sucide.Rate")]
png("heatmap.png", width=1200, height=1200)
cor_matrix <- cor(num_cols3)
col <- colorRampPalette(c("black", "darkred","red", "bisque1"))(200)
text_col <- ifelse(cor_matrix > 0.4 , "black", "white")
corrplot(cor_matrix, method="color",col = col ,addCoef.col=text_col, tl.col="black",
         tl.pos="lt", tl.cex=1, tl.srt=45, tl.offset = 0.5, cl.pos='r', cl.cex=0.6, cl.ratio=0.1)
dev.off()
```


## PCA
```{r}
# Perform PCA on the dataset (without : Country, Year, Gender, Life.expectancy)
df_events <- df[c("Year", "Unemployment", "Infant.Mortality", "GDP", "GNI", "Clean.fuels.and.cooking.technologies", "Per.Capita", "Mortality.caused.by.road.traffic.injury", "Tuberculosis.Incidence", "DPT.Immunization", "HepB3.Immunization", "Measles.Immunization", "Hospital.beds", "Basic.sanitation.services", "Tuberculosis.treatment", "Urban.population", "Rural.population", "Non.communicable.Mortality", "Sucide.Rate")]

pc_df_events <- prcomp(df_events, center = TRUE, scale. = TRUE)
```

```{r}
# Plot the variance per PC
fviz_screeplot(pc_df_events, addlabels = TRUE, ylim = c(0, 40), ncp = 18) 
```

```{r}
# Calculate the proportion of exaplained variance (PEV) from the std values
pc_df_events_var <- pc_df_events$sdev^2
pc_df_events_PEV <- pc_df_events_var / sum(pc_df_events_var)
```

```{r}
cumulative_pev <- cumsum(pc_df_events_PEV)

plot(cumulative_pev,
     type = "b", 
     pch = 20,
     cex = 1.5, 
     col = "blue",
     ylim = c(0, 1),
     xlab = "Number of Components",
     ylab = "Cumulative PEV",
     main = "Cumulative Proportion of Explained Variance")

abline(h = 0.8, col = 'red', lty = 'dashed')
abline(h = seq(0, 1, by = 0.1), col = "lightgray", lty = "dotted") 
abline(v = seq(1, length(cumulative_pev), by = 1), col = "lightgray", lty = "dotted")
```
```{r}
# Getting and Inspection of the Loadings
pc_df_events_loadings <- pc_df_events$rotation
pc_df_events_loadings
```

```{r}
color_palette <- rainbow(n = nrow(pc_df_events_loadings))

png("pc_loadings_variables_colored.png", width = 1800, height = 1300, res = 100)

par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))

for (i in 1:6) {
  barplot(
    pc_df_events_loadings[, i], 
    main = paste("PC", i),
    col = color_palette, 
    border = "white", 
    ylab = "Loadings",
    ylim = c(-1, 1) 
  )
  axis(2, seq(-1, 1, 0.1))
}
dev.off()
```
```{r}
png("variables_colors_legend.png", width = 800, height = 800, res = 100)

plot(0, type="n", xlim=c(0, 1), ylim=c(0, nrow(pc_df_events_loadings)), xlab="", ylab="", axes=FALSE)

# Legend with colors and names
for (i in 1:nrow(pc_df_events_loadings)) {
  rect(0.2, i-0.5, 0.3, i+0.4, col=color_palette[i], border="white")
  text(0.4, i, labels=row.names(pc_df_events_loadings)[i], pos=4)
}
dev.off()
```

```{r}
# Generate a biplot for PC1 and PC2
png("Biplot.png", width=3000, height=3000)
biplot(
  pc_df_events,
  scale = 0,
  col = c('grey40','red')
)
dev.off()
```
## Merge the original data with the PCs
```{r}
# merging df with pc_df_events$x in one unified dataframe
df_pc <- cbind(df, pc_df_events$x[, 1:7])
head(df_pc)
```

## Visualize the correlation matrix of the original data with the PCs
```{r}
num_cols_pc <- df_pc[c("Year","Life.expectancy", "Unemployment", "Infant.Mortality", "GDP", "GNI", "Clean.fuels.and.cooking.technologies", "Per.Capita", "Mortality.caused.by.road.traffic.injury", "Tuberculosis.Incidence", "DPT.Immunization", "HepB3.Immunization", "Measles.Immunization", "Hospital.beds", "Basic.sanitation.services", "Tuberculosis.treatment", "Urban.population", "Rural.population", "Non.communicable.Mortality", "Sucide.Rate", "PC1", "PC2", "PC3", "PC4", "PC5", "PC6", "PC7")]

png("heatmap_pc.png", width=1200, height=1200)
cor_matrix_pc <- cor(num_cols_pc)
col <- colorRampPalette(c("black", "darkred","red", "bisque1"))(200)
text_col <- ifelse(cor_matrix_pc > 0.4 , "black", "white")
corrplot(cor_matrix_pc, method="color",col = col ,addCoef.col=text_col, tl.col="black",
         tl.pos="lt", tl.cex=1, tl.srt=45, tl.offset = 0.5, cl.pos='r', cl.cex=0.6, cl.ratio=0.1)
dev.off()

```
##Clustering Analysis with hierarchical clustering
```{r}
# Scale the data
num_cols3 <- df[c("Year","Life.expectancy", "Unemployment", "Infant.Mortality", "GDP", "GNI", "Clean.fuels.and.cooking.technologies", "Per.Capita", "Mortality.caused.by.road.traffic.injury", "Tuberculosis.Incidence", "DPT.Immunization", "HepB3.Immunization", "Measles.Immunization", "Hospital.beds", "Basic.sanitation.services", "Tuberculosis.treatment", "Urban.population", "Rural.population", "Non.communicable.Mortality", "Sucide.Rate")]

# Standardize the data
num_cols3_std <- scale(num_cols3)
```

```{r}
# Calculate the distance
dist_num_cols3 <- dist(num_cols3_std, method = "euclidean")
```

```{r}
# Hierarchical clustering algorithm

hc_num_cols3 <- hclust(dist_num_cols3, method = "complete")
hc_num_cols3

num_cols3.cluster <- cutree(hc_num_cols3, k = 6)

```
```{r}
# Plot the dendrogram

png("dendrogram.png", width= 6000, height=3000, res = 100)
plot(hc_num_cols3, cex = 0.6, hang = -1)
dev.off()
```
## Clustering Analysis with K-means clustering
```{r}
# Scale the data
num_cols_cl <- df[c("Year","Life.expectancy", "Unemployment", "Infant.Mortality", "GDP", "GNI", "Clean.fuels.and.cooking.technologies", "Per.Capita", "Mortality.caused.by.road.traffic.injury", "Tuberculosis.Incidence", "DPT.Immunization", "HepB3.Immunization", "Measles.Immunization", "Hospital.beds", "Basic.sanitation.services", "Tuberculosis.treatment", "Urban.population", "Rural.population", "Non.communicable.Mortality", "Sucide.Rate")]

# Standardize the data
num_cols_std <- scale(num_cols_cl)
```

```{r}
# Define a range of k values to test
k_values <- 2:10  # Example range of k values from 2 to 10

# Initialize a vector to store silhouette scores for each k
silhouette_scores <- numeric(length(k_values))

# Iterate over each k value
for (i in seq_along(k_values)) {
  k <- k_values[i]
  
  # Perform k-means clustering
  set.seed(123)  # for reproducibility
  kmeans_result <- kmeans(num_cols_std, centers = k)
  
  # Calculate silhouette score
  silhouette_object <- silhouette(kmeans_result$cluster, dist(num_cols_std))
  silhouette_scores[i] <- mean(silhouette_object[, "sil_width"])
}

# Visualize the silhouette scores for different k values
plot(k_values, silhouette_scores, type = "b", pch = 16, xlab = "Number of Clusters (k)", ylab = "Silhouette Score", main = "Silhouette Score for K-Means Clustering")

```
K-means clustering suggests that the optimal number of clusters is 2,where the silhoutte score is the highest. 

```{r}
png("Cluster121233.png", width=10000, height=10000, res=600)

# Perform k-means clustering with k=2 
set.seed(123)  # for reproducibility
kmeans_result <- kmeans(num_cols_std, centers = 2)

# Calculating silhouette score
silhouette_object <- silhouette(kmeans_result$cluster, dist(num_cols_std))
silhouette_scores <- mean(silhouette_object[, "sil_width"])

# silhouette score
print(silhouette_scores)

cluster_colors <- c("blue", "red")
plot(silhouette_object, col = cluster_colors, border = cluster_colors, main = "Silhouette Plot")
dev.off()
```

```{r}
# Visualize the clusters and save the plot as a PNG file
png("k means clusters.png", width=2500, height=2500)
fviz_cluster(list(data = num_cols_std, cluster = kmeans_result$cluster))
dev.off()
```

## Modeling Random Forest
```{r}
# install the tree and randomForest packages from CRAN
if(require(randomForest) == FALSE){
  install.packages('randomForest')
  library(randomForest)
}
```
## We split the data into training testing and validation sets
```{r}
set.seed(42)

n_rows <- nrow(df) 

# Calculate sizes for splits
n_train <-round(n_rows * 0.7)
n_test <- round(n_rows * 0.2)

# Sample indices for training set
training_idx <- sample(seq_len(n_rows), size=n_train)

# Sample indices for test set from the remaining indices
remaining_idx <- setdiff(seq_len(n_rows), training_idx)
test_idx <- sample(remaining_idx, size=n_test)

# The remaining indices will be used for the validation set
validation_idx <- setdiff(remaining_idx, test_idx)

# Split the dataframe according to the sampled indices
training_df <-df[training_idx, ]
testing_df <- df[test_idx, ]
validation_df <-df[validation_idx, ]
```

### Before PCA ###
```{r}
# define a formula for predicting Life expectancy
df_formula = Life.expectancy ~ Year + Unemployment + Infant.Mortality + GDP + GNI + Clean.fuels.and.cooking.technologies + Per.Capita + Mortality.caused.by.road.traffic.injury + Tuberculosis.Incidence + DPT.Immunization + HepB3.Immunization + Measles.Immunization +  Hospital.beds + Basic.sanitation.services + Tuberculosis.treatment + Urban.population + Rural.population + Non.communicable.Mortality + Sucide.Rate


rf_df <- randomForest(df_formula, ntree = 800, importance = T, data = training_df, mtry = 6)
rf_df
```

```{r}
# plot the variable importance according to the
varImpPlot(rf_df, type = 1)
```
```{r}
# compute the prediction for the random forest model
rf_df_pred <- predict(rf_df, testing_df[, c("Year", "Unemployment", "Infant.Mortality", "GDP", "GNI", "Clean.fuels.and.cooking.technologies", "Per.Capita", "Mortality.caused.by.road.traffic.injury", "Tuberculosis.Incidence", "DPT.Immunization", "HepB3.Immunization", "Measles.Immunization", "Hospital.beds", "Basic.sanitation.services", "Tuberculosis.treatment", "Urban.population", "Rural.population", "Non.communicable.Mortality", "Sucide.Rate")], type= "response")
```

```{r}
# Calculate the time taken for training and prediction
time_train <- system.time({
  rf_df <- randomForest(df_formula, ntree = 800, importance = T, data = training_df, mtry = 6)
})

time_predict <- system.time({
  rf_df_pred <- predict(rf_df, testing_df[, c("Year", "Unemployment", "Infant.Mortality", "GDP", "GNI", "Clean.fuels.and.cooking.technologies", "Per.Capita", "Mortality.caused.by.road.traffic.injury", "Tuberculosis.Incidence", "DPT.Immunization", "HepB3.Immunization", "Measles.Immunization", "Hospital.beds", "Basic.sanitation.services", "Tuberculosis.treatment", "Urban.population", "Rural.population", "Non.communicable.Mortality", "Sucide.Rate")], type= "response")
})

print(time_train)
print(time_predict)
```

## Evaluate the model
```{r}
actual_values <- testing_df$Life.expectancy
#MSE
mse <- mean((rf_df_pred - testing_df$Life.expectancy)^2)
print(paste("MSE:", mse))

#RMSE
rmse <- sqrt(mse)
print(paste("RMSE:", rmse))

#MAE
mae <- mean(abs(rf_df_pred - testing_df$Life.expectancy))
print(paste("MAE:", mae))

#R-squared
rsq <- 1 - sum((actual_values - rf_df_pred)^2) / sum((actual_values - mean(actual_values))^2)
print(paste("R-squared:", rsq))
```


```{r}
df_comparison <- data.frame(Actual = actual_values, Predicted = rf_df_pred)
head(df_comparison)
```

```{r}
data <- data.frame(Actual = actual_values, Predicted = rf_df_pred)

ggplot(data, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.5) +  
  geom_smooth(method = "lm", color = "red", se = FALSE) + 
  labs(title = "Actual vs Predicted", x = "Actual values", y = "Predicted values") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5), text = element_text(size = 12))
```

### After PCA
```{r}
set.seed(42)

n_rows_pc <- nrow(df_pc) 

# Calculate sizes for splits
n_train_pc <-round(n_rows_pc * 0.7)
n_test_pc <- round(n_rows_pc * 0.2)

# Sample indices for training set
training_idx <- sample(seq_len(n_rows_pc), size=n_train_pc)

# Sample indices for test set from the remaining indices
remaining_idx <- setdiff(seq_len(n_rows_pc), training_idx)
test_idx <- sample(remaining_idx, size=n_test_pc)

# The remaining indices will be used for the validation set
validation_idx <- setdiff(remaining_idx, test_idx)

# Split the dataframe according to the sampled indices
pc_training <- df_pc[training_idx, ]
pc_testing <- df_pc[test_idx, ]
pc_validation <- df_pc[validation_idx, ]
```

```{r}
# define a formula for predicting 
df_formula_pc = Life.expectancy ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7

rf_df_pc <- randomForest(df_formula_pc, ntree = 300, importance = T, data = pc_training, mtry = 2)
rf_df_pc
```
```{r}
# plot the variable importancea according to the random forest model
varImpPlot(rf_df_pc, type = 1)
```
```{r}
# compute the prediction for the random forest model

rf_df_pc_pred <- predict(rf_df_pc, pc_testing[, c("PC1", "PC2", "PC3", "PC4", "PC5", "PC6", "PC7")], type= "response")

```

```{r}
# Calculate the time taken for training and prediction
time_train_pc <- system.time({
  rf_df_pc <- randomForest(df_formula_pc, ntree = 300, importance = T, data = pc_training, mtry = 2)
})

time_predict_pc <- system.time({
  rf_df_pred <- predict(rf_df_pc, pc_testing[, c("PC1", "PC2", "PC3", "PC4", "PC5", "PC6", "PC7")], type= "response")
})

print(time_train_pc)
print(time_predict_pc)
```

```{r}
#MSE
mse <- mean((rf_df_pc_pred - pc_testing$Life.expectancy)^2)
print(paste("MSE:", mse))

#RMSE
rmse <- sqrt(mse)
print(paste("RMSE:", rmse))

#MAE
mae <- mean(abs(rf_df_pc_pred - pc_testing$Life.expectancy))
print(paste("MAE:", mae))

#R-squared
rsq_pc <- 1 - sum((actual_values - rf_df_pc_pred)^2) / sum((actual_values - mean(actual_values))^2)
print(paste("R-squared:", rsq_pc))
```
```{r}
data_pc <- data.frame(Actual = actual_values, Predicted = rf_df_pc_pred)

ggplot(data_pc, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.5) +  
  geom_smooth(method = "lm", color = "red", se = FALSE) + 
  labs(title = "Actual vs Predicted", x = "Actual values", y = "Predicted values") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5), text = element_text(size = 12))
```

